---
title: "Investigating Changes in Whale Shark Abundnace in the Gulf of Mexico Over Time"
author: "Ana Bishop, Jasmime Papas, Tasneem Ahsanullah"
date: "2023-05-03"
output:
  html_document: default
  pdf_document: default
editor_options: 
  chunk_output_type: console
---
## Github repo link: 
https://github.com/anabishop99/AhsanullahBishopPapas_ENV872_EDA_FinalProject

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
```

```{r load libraries}
library(tidyverse)
library(lubridate)
library(trend)
library(zoo)
library(tseries)
library(dplyr)
library(cowplot)
library(trend)
library(Kendall)
library(here)
library(sf)
library(mapview); mapviewOptions(fgb = FALSE)
library(leaflet)
```

## Rationale

The whale shark species has been split into two major population groups, the Indo-Pacific Ocean and the Atlantic Ocean population, that make up 70% of the species total population. For our analysis, we decided to explore the Atlantic population, specifically in the Gulf of Mexico. The Gulf is a hot spot for feeding and migration and so we hoped to examine the whale shark abundance and its changes here.

First, we analyzed the whale shark observations in the Gulf of Mexico across a 7-year time frame to see if there were any major changes in abundance. We hypothesized that the abundance locations would vary by month depending on migration patterns.

Next, we were curious to see if climate change and rising sea temperatures could have an effect on the abundance of the whale sharks. The whale sharks generally prefer a temperature range of 69-77°F, but could this range change as the sea temperatures rise? We thought that the data may show a shift in location of whale sharks in order to stay in their desired temperature range. 

## Dataset Information

The Whale Shark Abundance data used in this analysis is from OBIS-SEAMAP. This was one of the only datasets that had multiple records over a long time period which is why we chose to use it. The whale shark data set spans from 1964 to 2010 and has 8,408 records of whale shark sightings. These sightings include chance encounters and individual catalogues. The encounters and photographs were mainly collected by the general public but the database is maintained by professional scientists.

The Sea Surface Temperature data is from the National Oceanic and Atmospheric Administration (NOAA). They have data on the global annual and monthly measure of sea surface temperature anomaly which is how different the temperature is from the average sea surface temperature. The time frame is from 1854 to present day and the data comes from the International Comprehensive Ocean-Atmosphere Dataset (ICODAS). We chose to download monthly data between the latitudes of 20N and 90N because that includes the Gulf of Mexico which is what we are interested in.

| Dataset Name                      | Data Source | Data Range   | Units                  | Variables used                 |
|---------------|---------------|---------------|---------------|---------------|
| Whale Shark Abundance             | OBIS-SEAMAP | 1964-2010    | Number of whale sharks | latitude, longitude, date-time |
| Sea Surface Temperature (Anomaly) | NOAA        | 1854-Present | Kelvin                 | year, month, anomaly           |

: Table 1: Information on Datasets


## Data Wrangling

First the whale shark dataset was filtered by latitude (17 to 30) and longitude (-98 to -79) to only include observations from the Gulf of Mexico. Date and time were separated into different columns and lubridate was used to turn the date into a date class. For the time series and linear regression analyses, we needed the count of whale sharks found each month. This was done by creating a column for just month and year. Then, a new column was created called total_sightings which was grouped by the month-year column. Lastly, the "distinct" function was used to only include the month-year column and sightings once for each month instead of for each day. Since, there is missing data points for certain months, we created a new data set of all the months between August 2002 and December 2009. This data set was then joined to the whale shark data set. All the months that didn't have data were NA so for the time series the NAs were replaced with zeros.

The sea surface temperature data was wrangling by removing unnecessary columns and the time range was filtered to only include data between August 2002 and December 2009. A month-year column was made similar to the column created for the whale shark data. The sea surface temperature data was combined (using left_join) with the whale shark data using the month-year column and all NAs were omitted.

```{r initial timeseries visulatization}
#read in data
gom_sightings <- read_csv("./Data/whaleshark_data/cleaned_gom_data.csv")

## calculate number of sightings per month/year
#isolate month and year
gom_sightings <- mutate(gom_sightings, month = month(date)) %>% 
  mutate(gom_sightings, year = year(date)) %>% 
  mutate(month_yr = my(paste(month,"-",year)))

#sum number of sightings per month-year
gom_sightings <- gom_sightings %>% group_by(month_yr) %>% mutate(total_sightings = n())

#need to fill in all missing months/years
#generate sequence of all months and years between August 2002 and December 2009
dates <- as.data.frame(seq(as.Date("2002-08-01"), as.Date("2009-12-01"), "months"))
colnames(dates)[1] ="month_yr"

#Select month/yr and total sightings from gom_sightings 
sightings <- gom_sightings %>% select(month_yr, total_sightings)
sightings <- distinct(sightings)
 
#left join that to the new sequence of months and years
sightings_all_months <- left_join(dates, sightings, by ="month_yr")

#Fill in all NAs as a total sighting of 0
sightings_all_months[is.na(sightings_all_months)] <- 0
```

```{r sst data, results='hide'}
#reading in raw sst data
sst_raw <- read.csv(here("Data/sst_correct_raw.csv"))

#Deleting empty columns
sst_raw_edit <- subset(sst_raw, select = -c(X, X.1, X.2))

#class is integer
class(sst_raw_edit$Year)

#creating a date column with month and year
sst_clean <- sst_raw_edit %>% 
  mutate(month_yr = my(paste(Month,"-",Year)))

#deleting first 7 rows so matches whale shark data
sst_clean <- sst_clean[-(1:7),]

#date class
class(sst_clean$month_yr)
```

```{r whale shark wrangling, echo=FALSE, include=FALSE}
#restricting data to gom
whale_points <- read.csv("./Data/whaleshark_data/dataset_points.csv", stringsAsFactors = T)

citation <- read_csv("./Data/whaleshark_data/dataset_citation.csv")

gom <- whale_points %>% filter(latitude >= 17 & latitude <= 30) %>% filter (longitude >= -98 & longitude <= -79)

gom_clean <- gom %>% separate(date_time, into = c('date','time'), sep=" ")
gom_clean$date <- ymd(gom_clean$date)
class(gom_clean$date)

#Separate month and year into their own columns
ws_count <- mutate(gom_clean, month = month(date)) %>% 
  mutate(gom_clean, year = year(date)) %>% 
  mutate(month_yr = my(paste(month,"-",year)))

#number of sightings per month-year
ws_count <- ws_count %>% group_by(month_yr) %>% mutate(total_sightings = n())

#Select month/yr and total sightings from gom_sightings 
sightings <- ws_count %>% select(month_yr, total_sightings)
sightings <- distinct(sightings)

#need to fill in all missing months/years
#generate sequence of all months and years between August 2002 and December 2009
dates <- as.data.frame(seq(as.Date("2002-08-01"), as.Date("2009-12-01"), "months"))
colnames(dates)[1] ="month_yr"

#left join that to the new sequence of months and years
sightings_all_months <- left_join(dates, sightings, by ="month_yr")

#Fill in all NAs as a total sighting of 0
sightings_all_months[is.na(sightings_all_months)] <- 0
```

```{r combining.data, echo=FALSE, include=FALSE}
#combining sst and whale shark sighting data
combined_clean <- left_join(sightings_all_months, sst_clean, by ="month_yr")
combined_clean <- na.omit(combined_clean)
```

```{r upper/lower lat analysis}
# latitudes: range from 20-29; break in half at 25 and compare two categories since there are a
# ton of obvs at lat = 21 and 27+
lower_lats <- gom_sightings %>% filter(latitude >= 20 & latitude < 25) #n = 241
upper_lats <- gom_sightings %>% filter(latitude >= 25 & latitude <= 30) #n = 40

## lower latitude ts
lower_lat_dates <- as.data.frame(seq(as.Date("2002-08-01"), as.Date("2009-12-01"), "months"))
colnames(lower_lat_dates)[1] ="month_yr"

#Select month/yr and total sightings from gom_sightings 
lower_lat_sightings <- lower_lats %>% select(month_yr, total_sightings)
lower_lat_sightings <- distinct(lower_lat_sightings)

#left join that to the new sequence of months and years
lower_lat_sightings_all_months <- left_join(lower_lat_dates, lower_lat_sightings, by ="month_yr")

#Fill in all NAs as a total sighting of 0
lower_lat_sightings_all_months[is.na(lower_lat_sightings_all_months)] <- 0

lower_lat_plot <- ggplot(lower_lat_sightings_all_months, aes(x = month_yr, y = total_sightings)) +
  geom_line() +
  labs(x = "Date", y = "Total Monthly Sightings", title = "Total Monthly Sightings in the Southern Latitudes of the Gulf of Mexico") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5)) 

## upper latitude ts
upper_lat_dates <- as.data.frame(seq(as.Date("2003-10-01"), as.Date("2009-08-01"), "months"))
colnames(upper_lat_dates)[1] ="month_yr"

#Select month/yr and total sightings from gom_sightings 
upper_lat_sightings <- upper_lats %>% select(month_yr, total_sightings)
upper_lat_sightings <- distinct(upper_lat_sightings)

#left join that to the new sequence of months and years
upper_lat_sightings_all_months <- left_join(upper_lat_dates, upper_lat_sightings, by ="month_yr")

#Fill in all NAs as a total sighting of 0
upper_lat_sightings_all_months[is.na(upper_lat_sightings_all_months)] <- 0

upper_lat_plot <- ggplot(upper_lat_sightings_all_months, aes(x = month_yr, y = total_sightings)) +
  geom_line() +
  labs(x = "Date", y = "Total Monthly Sightings", title = "Total Monthly Sightings in the Northern Latitudes of the Gulf of Mexico") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))
```

```{r east/west long analysis}
# longitudes: range from -80--96; break in half at 89 and compare two categories

eastern_longs <- gom_sightings %>% filter(longitude > -86 & longitude <= -80) #n = 12
western_longs <- gom_sightings %>% filter(longitude >= -97 & longitude <= -86) #n = 269

## eastern longitude ts
eastern_long_dates <- as.data.frame(seq(as.Date("2004-10-01"), as.Date("2009-08-01"), "months"))
colnames(eastern_long_dates)[1] ="month_yr"

#Select month/yr and total sightings from gom_sightings 
eastern_long_sightings <- eastern_longs %>% select(month_yr, total_sightings)
eastern_long_sightings <- distinct(eastern_long_sightings)

#left join that to the new sequence of months and years
eastern_long_sightings_all_months <- left_join(eastern_long_dates, eastern_long_sightings, by ="month_yr")

#Fill in all NAs as a total sighting of 0
eastern_long_sightings_all_months[is.na(eastern_long_sightings_all_months)] <- 0

eastern_long_plot <- ggplot(eastern_long_sightings_all_months, aes(x = month_yr, y = total_sightings)) +
  geom_line() +
  labs(x = "Date", y = "Total Monthly Sightings", title = "Total Monthly Sightings in the Eastern Longitudes of the Gulf of Mexico") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5)) 

## western longitude ts
western_long_dates <- as.data.frame(seq(as.Date("2002-08-01"), as.Date("2009-12-01"), "months"))
colnames(western_long_dates)[1] ="month_yr"

#Select month/yr and total sightings from gom_sightings 
western_long_sightings <- western_longs %>% select(month_yr, total_sightings)
western_long_sightings <- distinct(western_long_sightings)

#left join that to the new sequence of months and years
western_long_sightings_all_months <- left_join(western_long_dates, western_long_sightings, by ="month_yr")

#Fill in all NAs as a total sighting of 0
western_long_sightings_all_months[is.na(western_long_sightings_all_months)] <- 0

western_long_plot <- ggplot(western_long_sightings_all_months, aes(x = month_yr, y = total_sightings)) +
  geom_line() +
  labs(x = "Date", y = "Total Monthly Sightings", title = "Total Monthly Sightings in the Western Longitudes of the Gulf of Mexico") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5)) 
```

```{r individual lat analysis}
lat_20 <- gom_sightings %>% filter(latitude >= 20 & latitude < 21) # n = 2
lat_21 <- gom_sightings %>% filter(latitude >= 21 & latitude < 22) # n = 236
lat_22 <- gom_sightings %>% filter(latitude >= 22 & latitude < 23) # n = 1
lat_23 <- gom_sightings %>% filter(latitude >= 23 & latitude < 24) # n = 0
lat_24 <- gom_sightings %>% filter(latitude >= 24 & latitude < 25) # n = 2
lat_25 <- gom_sightings %>% filter(latitude >= 25 & latitude < 26) # n = 3
lat_26 <- gom_sightings %>% filter(latitude >= 26 & latitude < 27) # n = 2
lat_27 <- gom_sightings %>% filter(latitude >= 27 & latitude < 28) # n = 12
lat_28 <- gom_sightings %>% filter(latitude >= 28 & latitude < 29) # n = 18
lat_29 <- gom_sightings %>% filter(latitude >= 29 & latitude < 30) # n = 5

## set same date range for all using min/max of all lats. come back to make sure this is okay
all_lat_dates <- as.data.frame(seq(as.Date("2002-08-01"), as.Date("2009-12-01"), "months"))
colnames(all_lat_dates)[1] ="month_yr"

## lat 20

#Select month/yr and total sightings from gom_sightings 
lat_20_sightings <- lat_20 %>% select(month_yr, total_sightings)
lat_20_sightings <- distinct(lat_20_sightings)

#left join that to the new sequence of months and years
lat_20_sightings_all_months <- left_join(all_lat_dates, lat_20_sightings, by ="month_yr")

#Fill in all NAs as a total sighting of 0
lat_20_sightings_all_months[is.na(lat_20_sightings_all_months)] <- 0

lat_20_plot <- ggplot(lat_20_sightings_all_months, aes(x = month_yr, y = total_sightings)) +
  geom_line() +
  labs(x = "Date", y = "Total Monthly Sightings", title = expression("Total Monthly Sightings Between 20°N and 21°N")) +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5)) 

## lat 21

#Select month/yr and total sightings from gom_sightings 
lat_21_sightings <- lat_21 %>% select(month_yr, total_sightings)
lat_21_sightings <- distinct(lat_21_sightings)

#left join that to the new sequence of months and years
lat_21_sightings_all_months <- left_join(all_lat_dates, lat_21_sightings, by ="month_yr")

#Fill in all NAs as a total sighting of 0
lat_21_sightings_all_months[is.na(lat_21_sightings_all_months)] <- 0

lat_21_plot <- ggplot(lat_21_sightings_all_months, aes(x = month_yr, y = total_sightings)) +
  geom_line() +
  labs(x = "Date", y = "Total Monthly Sightings", title = "Total Monthly Sightings Between 21°N and 22°N") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5)) 

## lat 22

#Select month/yr and total sightings from gom_sightings 
lat_22_sightings <- lat_22 %>% select(month_yr, total_sightings)
lat_22_sightings <- distinct(lat_22_sightings)

#left join that to the new sequence of months and years
lat_22_sightings_all_months <- left_join(all_lat_dates, lat_22_sightings, by ="month_yr")

#Fill in all NAs as a total sighting of 0
lat_22_sightings_all_months[is.na(lat_22_sightings_all_months)] <- 0

lat_22_plot <- ggplot(lat_22_sightings_all_months, aes(x = month_yr, y = total_sightings)) +
  geom_line() +
  labs(x = "Date", y = "Total Monthly Sightings", title = "Total Monthly Sightings Between 22°N and 23°N") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5)) 

## lat 23

#Select month/yr and total sightings from gom_sightings 
lat_23_sightings <- lat_23 %>% select(month_yr, total_sightings)
lat_23_sightings <- distinct(lat_23_sightings)

#left join that to the new sequence of months and years
lat_23_sightings_all_months <- left_join(all_lat_dates, lat_23_sightings, by ="month_yr")

#Fill in all NAs as a total sighting of 0
lat_23_sightings_all_months[is.na(lat_23_sightings_all_months)] <- 0

lat_23_plot <- ggplot(lat_23_sightings_all_months, aes(x = month_yr, y = total_sightings)) +
  geom_line() +
  labs(x = "Date", y = "Total Monthly Sightings", title = "Total Monthly Sightings Between 23°N and 24°N") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))

## lat 24

#Select month/yr and total sightings from gom_sightings 
lat_24_sightings <- lat_24 %>% select(month_yr, total_sightings)
lat_24_sightings <- distinct(lat_24_sightings)

#left join that to the new sequence of months and years
lat_24_sightings_all_months <- left_join(all_lat_dates, lat_24_sightings, by ="month_yr")

#Fill in all NAs as a total sighting of 0
lat_24_sightings_all_months[is.na(lat_24_sightings_all_months)] <- 0

lat_24_plot <- ggplot(lat_24_sightings_all_months, aes(x = month_yr, y = total_sightings)) +
  geom_line() +
  labs(x = "Date", y = "Total Monthly Sightings", title = "Total Monthly Sightings Between 24°N and 25°N") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5)) 

## lat 25

#Select month/yr and total sightings from gom_sightings 
lat_25_sightings <- lat_25 %>% select(month_yr, total_sightings)
lat_25_sightings <- distinct(lat_25_sightings)

#left join that to the new sequence of months and years
lat_25_sightings_all_months <- left_join(all_lat_dates, lat_25_sightings, by ="month_yr")

#Fill in all NAs as a total sighting of 0
lat_25_sightings_all_months[is.na(lat_25_sightings_all_months)] <- 0

lat_25_plot <- ggplot(lat_25_sightings_all_months, aes(x = month_yr, y = total_sightings)) +
  geom_line() +
  labs(x = "Date", y = "Total Monthly Sightings", title = "Total Monthly Sightings Between 25°N and 26°N") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5)) 

## lat 26

#Select month/yr and total sightings from gom_sightings 
lat_26_sightings <- lat_26 %>% select(month_yr, total_sightings)
lat_26_sightings <- distinct(lat_26_sightings)

#left join that to the new sequence of months and years
lat_26_sightings_all_months <- left_join(all_lat_dates, lat_26_sightings, by ="month_yr")

#Fill in all NAs as a total sighting of 0
lat_26_sightings_all_months[is.na(lat_26_sightings_all_months)] <- 0

lat_26_plot <- ggplot(lat_26_sightings_all_months, aes(x = month_yr, y = total_sightings)) +
  geom_line() +
  labs(x = "Date", y = "Total Monthly Sightings", title = "Total Monthly Sightings Between 26°N and 27°N") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5)) 

## lat 27

#Select month/yr and total sightings from gom_sightings 
lat_27_sightings <- lat_27 %>% select(month_yr, total_sightings)
lat_27_sightings <- distinct(lat_27_sightings)

#left join that to the new sequence of months and years
lat_27_sightings_all_months <- left_join(all_lat_dates, lat_27_sightings, by ="month_yr")

#Fill in all NAs as a total sighting of 0
lat_27_sightings_all_months[is.na(lat_27_sightings_all_months)] <- 0

lat_27_plot <- ggplot(lat_27_sightings_all_months, aes(x = month_yr, y = total_sightings)) +
  geom_line() +
  labs(x = "Date", y = "Total Monthly Sightings", title = "Total Monthly Sightings Between 27°N and 28°N") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))

## lat 28

#Select month/yr and total sightings from gom_sightings 
lat_28_sightings <- lat_28 %>% select(month_yr, total_sightings)
lat_28_sightings <- distinct(lat_28_sightings)

#left join that to the new sequence of months and years
lat_28_sightings_all_months <- left_join(all_lat_dates, lat_28_sightings, by ="month_yr")

#Fill in all NAs as a total sighting of 0
lat_28_sightings_all_months[is.na(lat_28_sightings_all_months)] <- 0

lat_28_plot <- ggplot(lat_28_sightings_all_months, aes(x = month_yr, y = total_sightings)) +
  geom_line() +
  labs(x = "Date", y = "Total Monthly Sightings", title = "Total Monthly Sightings Between 28°N and 29°N") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))

## lat 29

#Select month/yr and total sightings from gom_sightings 
lat_29_sightings <- lat_29 %>% select(month_yr, total_sightings)
lat_29_sightings <- distinct(lat_29_sightings)

#left join that to the new sequence of months and years
lat_29_sightings_all_months <- left_join(all_lat_dates, lat_29_sightings, by ="month_yr")

#Fill in all NAs as a total sighting of 0
lat_29_sightings_all_months[is.na(lat_29_sightings_all_months)] <- 0

lat_29_plot <- ggplot(lat_29_sightings_all_months, aes(x = month_yr, y = total_sightings)) +
  geom_line() +
  labs(x = "Date", y = "Total Monthly Sightings", title = "Total Monthly Sightings Between 29°N and 30°N") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))
```

```{r compare all individual longs}
# doing longs - counting 2 by 2
long_80 <- gom_sightings %>% filter(longitude < -80 & longitude >= -82) # n = 3
long_82 <- gom_sightings %>% filter(longitude < -82 & longitude >= -84) # n = 8
long_84 <- gom_sightings %>% filter(longitude < -84 & longitude >= -86) # n = 1
long_86 <- gom_sightings %>% filter(longitude < -86 & longitude >= -88) # n = 240
long_88 <- gom_sightings %>% filter(longitude < -88 & longitude >= -90) # n = 8
long_90 <- gom_sightings %>% filter(longitude < -90 & longitude >= -92) # n = 15
long_92 <- gom_sightings %>% filter(longitude < -92 & longitude >= -94) # n = 3
long_94 <- gom_sightings %>% filter(longitude < -94 & longitude >= -96) # n = 1
long_96 <- gom_sightings %>% filter(longitude < -96 & longitude >= -98) # n = 2

## set same date range for all using min/max of all lats. come back to make sure this is okay
all_long_dates <- as.data.frame(seq(as.Date("2002-08-01"), as.Date("2009-12-01"), "months"))
colnames(all_long_dates)[1] ="month_yr"

## long 80

#Select month/yr and total sightings from gom_sightings 
long_80_sightings <- long_80 %>% select(month_yr, total_sightings)
long_80_sightings <- distinct(long_80_sightings)

#left join that to the new sequence of months and years
long_80_sightings_all_months <- left_join(all_lat_dates, long_80_sightings, by ="month_yr")

#Fill in all NAs as a total sighting of 0
long_80_sightings_all_months[is.na(long_80_sightings_all_months)] <- 0

long_80_plot <- ggplot(long_80_sightings_all_months, aes(x = month_yr, y = total_sightings)) +
  geom_line() +
  labs(x = "Date", y = "Total Monthly Sightings", title = "Total Monthly Sightings Between 80°W and 82°W") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5)) 

## lat 21

#Select month/yr and total sightings from gom_sightings 
long_82_sightings <- long_82 %>% select(month_yr, total_sightings)
long_82_sightings <- distinct(long_82_sightings)

#left join that to the new sequence of months and years
long_82_sightings_all_months <- left_join(all_lat_dates, long_82_sightings, by ="month_yr")

#Fill in all NAs as a total sighting of 0
long_82_sightings_all_months[is.na(long_82_sightings_all_months)] <- 0

long_82_plot <- ggplot(long_82_sightings_all_months, aes(x = month_yr, y = total_sightings)) +
  geom_line() +
  labs(x = "Date", y = "Total Monthly Sightings", title = "Total Monthly Sightings Between 82°W and 84°W") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5)) 

## long 84

#Select month/yr and total sightings from gom_sightings 
long_84_sightings <- long_84 %>% select(month_yr, total_sightings)
long_84_sightings <- distinct(long_84_sightings)

#left join that to the new sequence of months and years
long_84_sightings_all_months <- left_join(all_lat_dates, long_84_sightings, by ="month_yr")

#Fill in all NAs as a total sighting of 0
long_84_sightings_all_months[is.na(long_84_sightings_all_months)] <- 0

long_84_plot <- ggplot(long_84_sightings_all_months, aes(x = month_yr, y = total_sightings)) +
  geom_line() +
  labs(x = "Date", y = "Total Monthly Sightings", title = "Total Monthly Sightings Between 84°W and 86°W") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5)) 

## long 86

#Select month/yr and total sightings from gom_sightings 
long_86_sightings <- long_86 %>% select(month_yr, total_sightings)
long_86_sightings <- distinct(long_86_sightings)

#left join that to the new sequence of months and years
long_86_sightings_all_months <- left_join(all_lat_dates, long_86_sightings, by ="month_yr")

#Fill in all NAs as a total sighting of 0
long_86_sightings_all_months[is.na(long_86_sightings_all_months)] <- 0

long_86_plot <- ggplot(long_86_sightings_all_months, aes(x = month_yr, y = total_sightings)) +
  geom_line() +
  labs(x = "Date", y = "Total Monthly Sightings", title = "Total Monthly Sightings Between 86°W and 88°W") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))

## long 88

#Select month/yr and total sightings from gom_sightings 
long_88_sightings <- long_88 %>% select(month_yr, total_sightings)
long_88_sightings <- distinct(long_88_sightings)

#left join that to the new sequence of months and years
long_88_sightings_all_months <- left_join(all_lat_dates, long_88_sightings, by ="month_yr")

#Fill in all NAs as a total sighting of 0
long_88_sightings_all_months[is.na(long_88_sightings_all_months)] <- 0

long_88_plot <- ggplot(long_88_sightings_all_months, aes(x = month_yr, y = total_sightings)) +
  geom_line() +
  labs(x = "Date", y = "Total Monthly Sightings", title = "Total Monthly Sightings Between 88°W and 90°W") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5)) 

## long 90

#Select month/yr and total sightings from gom_sightings 
long_90_sightings <- long_90 %>% select(month_yr, total_sightings)
long_90_sightings <- distinct(long_90_sightings)

#left join that to the new sequence of months and years
long_90_sightings_all_months <- left_join(all_lat_dates, long_90_sightings, by ="month_yr")

#Fill in all NAs as a total sighting of 0
long_90_sightings_all_months[is.na(long_90_sightings_all_months)] <- 0

long_90_plot <- ggplot(long_90_sightings_all_months, aes(x = month_yr, y = total_sightings)) +
  geom_line() +
  labs(x = "Date", y = "Total Monthly Sightings", title = "Total Monthly Sightings Between 90°W and 92°W") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5)) 

## long 92

#Select month/yr and total sightings from gom_sightings 
long_92_sightings <- long_92 %>% select(month_yr, total_sightings)
long_92_sightings <- distinct(long_92_sightings)

#left join that to the new sequence of months and years
long_92_sightings_all_months <- left_join(all_lat_dates, long_92_sightings, by ="month_yr")

#Fill in all NAs as a total sighting of 0
long_92_sightings_all_months[is.na(long_92_sightings_all_months)] <- 0

long_92_plot <- ggplot(long_92_sightings_all_months, aes(x = month_yr, y = total_sightings)) +
  geom_line() +
  labs(x = "Date", y = "Total Monthly Sightings", title = "Total Monthly Sightings Between 92°W and 94°W") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5)) 

## long 94

#Select month/yr and total sightings from gom_sightings 
long_94_sightings <- long_94 %>% select(month_yr, total_sightings)
long_94_sightings <- distinct(long_94_sightings)

#left join that to the new sequence of months and years
long_94_sightings_all_months <- left_join(all_lat_dates, long_94_sightings, by ="month_yr")

#Fill in all NAs as a total sighting of 0
long_94_sightings_all_months[is.na(long_94_sightings_all_months)] <- 0

long_94_plot <- ggplot(long_94_sightings_all_months, aes(x = month_yr, y = total_sightings)) +
  geom_line() +
  labs(x = "Date", y = "Total Monthly Sightings", title = "Total Monthly Sightings Between 94°W and 96°W") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))

## long 96

#Select month/yr and total sightings from gom_sightings 
long_96_sightings <- long_96 %>% select(month_yr, total_sightings)
long_96_sightings <- distinct(long_96_sightings)

#left join that to the new sequence of months and years
long_96_sightings_all_months <- left_join(all_lat_dates, long_96_sightings, by ="month_yr")

#Fill in all NAs as a total sighting of 0
long_96_sightings_all_months[is.na(long_96_sightings_all_months)] <- 0

long_96_plot <- ggplot(long_96_sightings_all_months, aes(x = month_yr, y = total_sightings)) +
  geom_line() +
  labs(x = "Date", y = "Total Monthly Sightings", title = "Total Monthly Sightings Between 96°W and 98°W") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))
```

```{r}
#removing unneccessary data
gom_final<- gom_clean %>% 
  select(dataset_id, row_id, latitude, longitude, species_name, date, time, last_mod)
```

## Exploratory Analysis
Upon beginning the analysis, our main questions were these: has the abundance of whale sharks in the Gulf of Mexico changed over time? If so, did they change in relation to warming sea surface temperatures (SST)? To find out, we first visualized the total sightings of whale sharks in the Gulf of Mexico across our entire study period (2002-2009; Figure XX). This visualization confirmed that a large increase in whale shark abundance appeared to occur after 2008, which gave us the confidence to proceed with our analysis. We then investigated how SST had changed in the Gulf over the course of our study period to confirm that the water temperature had indeed changed (Figure XX). Upon seeing that it had (check this!!), we knew that there was merit to our investigation, and proceeded to look at more nuanced questions within our study.

After seeing that there appeared to be an overall increase in whale shark abundance, we were curious to see if whale shark abundances had changed within the Gulf of Mexico. Was it possible that the warming waters were causing them to potentially migrate to the Gulf in larger numbers? If so, could the changing water have any impact on which region of the Gulf that they favored? We divided the study area into northern and southern halves, along with eastern and western halves, to see if there appeared to be any sort of trade off between the regions over time (Figure XX). The results showed that there was no clear trade off that could be seen. Instead, it appeared that abundance generally increased across all four regions over time. As one last attempt to view any sort of habitat usage trade off, we then compared the abundance of whale sharks over the study period between each latitude parallel of the Gulf of Mexico (Figure XX), along with their abundance between every two longitute parallels (Figure XX). Again, there was no distinct trade off in usage seen; instead, it again appeared that the whale shark abundance simply increased overall in almost every area. This confired that we should not conduct an analysis to see if habitat preferences had changed, and instead simply focus on whether the number of sighted whale sharks had significantly increased over time or not. 

``` {r time series plot}
#initial timeseries
ggplot(sightings_all_months, aes(x = month_yr, y = total_sightings)) +
  geom_line() +
  labs(x = "Date", y = "Total Monthly Sightings", title = "Total Monthly Sightings in the Gulf of Mexico From 2002-2009", caption = "Fig. 1. Time series of montly whale shark sightings in the Gulf of Mexico for the entire study period") +
  theme_minimal() +
  theme(plot.caption = element_text(hjust = 0)) +
  theme(plot.title = element_text(hjust = 0.5)) 
```

<br> 

```{r SST.plot, echo=FALSE, message=FALSE, warning=FALSE, fig.cap="Figure 2. Sea Surface Temperature over Time"}
sst.plot <-
  ggplot(sst_clean, aes(x=month_yr,y=Anomaly))+
  geom_point()+
  geom_smooth(method="lm",color="blue",se=FALSE)+
  labs(
  x="Date",
  y="Sea Surface Temperature (Anomaly)",
  title = "Sea Surface Temperature over Time"
  )

print(sst.plot)
```

Figure 2 shows that sea surface temperature anomaly is increasing slightly over time.

<br>

```{r compare lat/long analysis, fig.height=10}
# compare all plots together
plot_grid(upper_lat_plot, lower_lat_plot, eastern_long_plot, western_long_plot, nrow = 4, align = 'v', rel_heights = c(1, 1, 1, 1))
```

Figure 3. A four-panel comparison of whale shark sightings in the northern and southern halves of the Gulf of Mexico, followed by the Eastern and Western halves. Each category of sightings had different time periods associated with them, so note the differences in the x axes.

<br>

```{r comparison lat plot, fig.height= 20, fig.width= 10}
# compare all plots together
lat_comparison_plot <- plot_grid(lat_20_plot, lat_21_plot, lat_22_plot, lat_23_plot, lat_24_plot, lat_25_plot, lat_26_plot, lat_27_plot, lat_28_plot, lat_29_plot, nrow = 10, align = 'v', rel_heights = c(.1, .1, .1, .1, .1, .1, .1, .1, .1, .1))
print(lat_comparison_plot)
```

Figure 4. A comparison of whale shark sightings within each latitude parallel of the Gulf of Mexcio.

<br> 

```{r individual long plot comparison, fig.height= 20, fig.width= 10}
# compare all plots together
long_comparison_plot <- plot_grid(long_80_plot, long_82_plot, long_84_plot, long_86_plot, long_88_plot, long_90_plot, long_92_plot, long_94_plot, long_96_plot, nrow = 9, align = 'v', rel_heights = c(.1, .1, .1, .1, .1, .1, .1, .1, .1))
print(long_comparison_plot)
```

Figure 5. A comparison of whale shark sightings within every two longitude parallels of the Gulf of Mexico.

```{r}
gom_sf<-st_as_sf(gom_final, coords = c("longitude","latitude"), crs=4326)

#plot of map
#ggplot(gom_sf)+
 # geom_sf(aes(geometry=geometry))

#map projected
#mapview(gom_sf)

#Monthly Map
gom_final<- mutate(gom_sf, month= month(date))
gom_final<- mutate(gom_final, year=year(date))


mapview(gom_final, zcol= "month", at = seq(1:12), legend = TRUE)

```
Figure 6. This map shows the location of whale shark observations by month

As you can see, the whale sharks seem to be aggregating around the Yucatan Peninsula during the late summer/early fall months and migrating to the northern region of the Gulf during the winter months. 

```{r}
gom_sft<-st_as_sf(gom, coords = c("longitude","latitude"), crs=4326)
#mapview(gom_sft)
```

## Analysis

### Time Series Analysis of Whale Shark Sightings Over Time

```{r monthly time series, echo=FALSE, include=FALSE, message=FALSE, warning=FALSE}

##### Next Analysis - test for increasing trend over all time. Need to take out seasonality ####
# set time series object
gom_ts <- ts(sightings_all_months$total_sightings, start = c(2002,8), frequency = 12)

# decompose time series
gom_ts_decomposed <- stl(gom_ts, s.window = "periodic")

# run monotonic trend analysis - seasonal Mann Kendall ##### NOTE! This had too few observations to be valid. I'll do it by year instead
gom_sighting_analysis <- Kendall::SeasonalMannKendall(gom_ts)

#show results
gom_sighting_analysis
summary(gom_sighting_analysis)
```

```{r, echo=FALSE, warning=FALSE, fig.cap="Figure 7. The decomposed time series of monthly sightings in the Gulf of Mexico"}
plot(gom_ts_decomposed)
```

The increasing trend of monthly sightings in the Gulf of Mexico was significant (tau = 0.32, p = 0.002). However, there were not enough sightings per month for the analysis to be valid. Therefore, I aggregated the sightings by year instead and analyzed the time series again.

```{r yearly time series, echo=FALSE, include=FALSE, message=FALSE, warning=FALSE}
### new attempt, grouping by year
gom_yearly_sightings <- gom_sightings %>% group_by(year) %>% mutate(total_sightings = n())

# all years are already filled in

#Select month/yr and total sightings from gom_sightings 
yearly_sightings <- gom_yearly_sightings %>% select(year, total_sightings)
yearly_sightings <- distinct(yearly_sightings)

# set time series object
gom_yearly_ts <- ts(yearly_sightings$total_sightings, start = c(2002), frequency = 1)

# decompose time series
# gom_yearly_ts_decomposed <- stl(gom_yearly_ts) - won't work

# plot(gom_yearly_ts_decomposed) #won't work bc of previous error

# run monotonic trend analysis -  Mann Kendall and lm ##### NOTE! This had too few observations to be valid. I'll do it by year instead
gom_yearly_sighting_analysis_mk <- Kendall::MannKendall(gom_yearly_ts)
gom_yearly_sighting_analysis_lm <- lm(year ~ total_sightings, data = yearly_sightings)

#show results
gom_yearly_sighting_analysis_mk
summary(gom_yearly_sighting_analysis_mk)

gom_yearly_sighting_analysis_lm
summary(gom_yearly_sighting_analysis_lm)
```

```{r, echo= FALSE, message = FALSE}
#initial timeseries
ggplot(yearly_sightings, aes(x = year, y = total_sightings)) +
  geom_line() +
  labs(x = "Date", y = "Total Yearly Sightings", title = "Total Yearly Sightings in the Gulf of Mexico", caption = "Fig. 8. Total sightings in the Gulf of Mexico for the entire study period, aggregated by year") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5)) +
  theme(plot.caption = element_text(hjust = 0))
```

The results of this analysis did not show that there was an increasing trend using a Mann-Kendall analysis (tau = 0.43, p = 0.17). However, we conducted a linear regression analysis on the data as well, and that test did return positive results (p = 0.04). However, due to the conflicting results of the tests and the p value from the linear regression still being rather large, we felt that a third analysis could improve the strength of our conclusions. Therefore, we isolated the months in which whale sharks were the most frequently seen (June - September; Figure 8), and performed a final Mann-Kendall analysis on those months. The results showed that there was a significant increase in sightings over those "hot spot" months over time (tau = 0.38, p = 0.004).

```{r seasonal time series, echo=FALSE, include=FALSE, message=FALSE, warning=FALSE}
## aggregate by high sighting months only then do a mann-kendall - NOT SEASONAL - on that!

# analysis on only June-Sept
summer_fall <- gom_sightings %>% filter(month %in% 6:9)
summer_fall_sightings <- summer_fall %>% select(month_yr, total_sightings)
summer_fall_sightings <- distinct(summer_fall_sightings)

# generate missing values
summer_fall_dates <- as.data.frame(seq(as.Date("2002-08-01"), as.Date("2009-12-01"), "months"))
colnames(summer_fall_dates)[1] ="month_yr"
summer_fall_dates <- summer_fall_dates %>% filter(month(month_yr) %in% 6:9)

#left join that to the new sequence of months and years
summer_fall_sightings <- left_join(summer_fall_dates, summer_fall_sightings, by ="month_yr")

#Fill in all NAs as a total sighting of 0
summer_fall_sightings[is.na(summer_fall_sightings)] <- 0

# set time series object
summer_fall_ts <- ts(summer_fall_sightings$total_sightings, start = c(2002), frequency = 1)

# decompose time series
# summer_fall_ts_decomposed <- stl(summer_fall_ts) # not working

# plot(gom_yearly_ts_decomposed) #won't work bc of previous error

# run monotonic trend analysis -  Mann Kendall
summer_fall_analysis_mk <- Kendall::MannKendall(summer_fall_ts)

#show results
summer_fall_analysis_mk # significant!
```

```{r, fig.cap="Fig. 9. Time series of whale shark sightings over the months in which whale sharks were most frequently spotted (June-September)"}
ggplot(summer_fall_sightings, aes(x = month_yr, y = total_sightings)) +
  geom_line() +
  labs(x = "Date", y = "Total Monthly Sightings", title = "Total Monthly Sightings from June through September") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))
```

Map of Monthly Hotspots
```{r}
#monthly data
monthly<- gom_final %>% 
  filter(month==6|month== 7| month== 8| month==9)

#Fig 10. This map portrays the location of whale shark observations during the months of June through September as these months seem to have the most sightings. 

#map of monthly hotspots
mapview(monthly, zcol= "month", cex= 3)

#map of monthly hotspot by year
monthly2002<- monthly %>% 
  filter(year==2002)
mapview(monthly2002, zcol= "month")

monthly2003<- monthly %>% 
  filter(year==2003)
mapview(monthly2003, zcol= "month")

monthly2004<- monthly %>% 
  filter(year==2004)
mapview(monthly2004, zcol= "month")
```
Caption: We decided mapping the whale shark abundance by year was not the most effective way to observe the data as there were too many years that had too little observations or only had observations for a single month. 

Whale Shark data by Year
```{r}
mapview(gom_final, zcol= "year")
```
Caption: We encourage you to zoom in to the tip of the Yucatan Penninsula to better observe the abundance of whale sharks in this region. As you can see, there seems to be more whale sharks moving south towards the equator during 2009. 

Observations in June By Year
```{r}
june<- gom_final %>% 
  filter(month==6)
mapview(june, zcol= "year")
```

Observations in July By Year
```{r}
july<- gom_final %>% 
  filter(month==7)
mapview(july, zcol= "year")
```

Observations in August By Year
```{r}
august<- gom_final %>% 
  filter(month==8)
mapview(august, zcol= "year")
```

Observations In September By Year
```{r}
september<- gom_final %>% 
  filter(month==9)
mapview(september, zcol= "year")
```

### Sea Surface Temperature vs. Whale Shark Sightings

```{r sst.whale, echo=FALSE, message=FALSE, warning=FALSE, fig.cap="Figure 9. Sea Surface Temperature vs. Whale Shark Sightings"}
#plot of whale shark sightings by sea surface temperature anomalies
combined.plot <-
  ggplot(combined_clean, aes(x=Anomaly,y=total_sightings))+
  geom_point()+
  geom_smooth(method="lm",color="blue",se=FALSE)+
  labs(
  x="Sea Surface Temperature (Anomaly)",
  y="Whale Shark Sightings",
  title = "Sea Surface Temperature vs. Whale Shark Sightings"
  )

print(combined.plot)
```

Figure 9 shows that there is an increase in whale shark sightings as sea surface temperature anomaly increases. To analyze this increase we used a GLM. The results of the GLM showed that this increase is not significant (p-value = 0.4093, DF=36, R-squared= 0.019).

```{r lm, echo=FALSE, include=FALSE, message=FALSE, warning=FALSE}
#linear regression of whale shark sightings by sst
combined.regression <- lm(data=combined_clean,total_sightings~Anomaly)
summary(combined.regression)
```

## Summary

Overall, there was a significant increase in whale shark sightings overtime throughout the study area. While this could be an encouraging result that may indicate the whale shark population might be growing, and/or may be using the Gulf of Mexico in higher numbers, there are other confounding variables at play. Given the opportunistic nature of this dataset, it's possible that there simply may have been more people looking for whale sharks in the later years of the study period. This could be because the whales were in a more favorable environment, or maybe the public was better educated about how to submit sightings to the database managers. In summary, while there was a signficiant increasing trend in sightings over the years, we cannot be sure about how or why that occurred. Conducting systematic surveys looking for whale sharks within the Gulf that have a standardized methodology in the future could help give more confidence in the findings of this analysis, and indicate if there actually are more whales visiting the Gulf of Mexico.

Sea surface temperature did not affect whale shark sightings significantly. This could be because the data we looked at only spanned a few years or because the whale shark data was opportunistic so there may have been more whale sharks than counted.


